An Evolutionary Algorithm for the Multiobjective Risk-Equity Constrained Routing Problem

### 2011-11-30 trunk/DAW_hazmatHypE/trunk/paper/hazmatRouting_EvolOptim.tex
final revision of manuscript including spell-check

\begin{abstract}
Nowadays, various hazardous materials, or \emph{hazmats}, are transported on our streets, via trains, planes or ships. In paricular when routing them over the street network, where typically several possible routes exist, one would like to minimize both the costs and the total risks. However, sometimes one aims at minimizing an additional type of risk, which in most optimization models is resulting in a non-linear objective: the equity risk, or the \emph{maximum} risk over all regions, streets, etc. In other words, one aims at distributing the overall risks as much as possible while the costs and the total risk should be still minimized simultaneously. In this paper, this three-objective optimization problem is tackled with a novel randomized algorithm, based on a state-of-the-art Evolutionary Multiobjective Optimization algorithm and taylored to the needs of finding Pareto-optimal truck routes in a given network by proposing a variable-length representation and problem-specific operators. Experimental results on realistic instances for the street network of Italy's Lazio region show that the proposed approach efficiently finds a variety of good solutions in reasonable time. The experiments also show the influence of the algorithm's different building blocks and the instances' parameters on the solution quality.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The transportation of hazardous materials ({\em hazmat} from now on) has received a large interest in recent years, which results from the increase in public awareness of the dangers of hazmats and the enormous amount of hazmats being transported \citep{etv2007a,cd2008a}. The main target of this problem is to select routes from a given origin-destination pair of nodes such that the risk for the surrounding population and the environment is minimized---without producing excessive economic costs.

More specifically, one can distinguish between two main problems of different scale \citep{etv2007a}. The \emph{local} hazmat routing problem aims at finding the route for a single shipment from a source to a destination node in a given network; the \emph{global} hazmat routing problem, on the other hand, looks as the macro level of shipping several hazmat types simultaneously from multiple sources to multiple destinations---with much less studies on the latter \citep{etc2007a}. 
When solving such a global hazmat routing problem by minimizing both cost and the total risk, typically several vehicles share the same (short) routes which results in high risks associated to regions surrounding these paths whereas other regions might not be affected at all. In this case, one may wish to distribute the risk in an equitable way over the population and the environment. Although several studies consider this additional minimization of the equity risk, most of them restrict the model to a \emph{single} origin-destination hazmat routing for a specific hazmat, transport mode and vehicle type (see for example \cite{aeb2000a}).
More realistic \emph{multi}-commodity flow model have been proposed where each commodity is considered as one hazmat type \citep{cd2008a} but the majority of all hazmat routing studies do not consider the multi-objective nature of the hazmat routing problem but optimize only the risk \cite{gkbk1990a} or an aggregation or scalarization of the objective functions instead \citep{cd2008a}. An exception is \citep{dgs2005a} where the focus lies on finding a set of dissimilar paths that are Pareto-optimal with respect to cost and risk instead of solving the multiobjective hazmat routing problem directly. Moreover, only one origin-destination pair is considered.

As it is important for a decision maker to study the trade-offs among the objectives, we treat the global hazmat routing problem of finding a set of routes minimizing total costs, total risk, and equity risk, from a multiobjective perspective. Evolutionary Multiobjective Optimization (EMO) algorithms have been shown to be able to compute a set of solutions showing the trade-offs between multiple objectives within a single algorithm run in particular when other (exact) algorithms have difficulties to approach the Pareto front, e.g, if the front is too large. As general black-box algorithms, they also allow to cope with changes in the problem formulation which occurs frequently in real-world applications.

In this study, we therefore propose an EMO algorithm for the global hazmat routing problem (Sec.~\ref{S_MEO}). Before, we formalize the specific multiobjective risk-equity constrained routing problem with three objectives (minimize total routing cost, total routing risk {\em and} risk equity) as a multicommodity flow model as in \cite{cd2008a} since this model is the most realistic one permitting to manage several hazmat types simultaneously (Sec.~\ref{S_RECRP}). Section~\ref{sec:experiments} is then devoted to the experimental analysis of this algorithm and its application to realistic instances in the Lazio region of Italy before Sec.~\ref{S_FW} concludes the paper.

@@

Let the transportation network be represented as a directed graph $G = (N, A)$, with $N$ being the set of nodes and $A$ the set of arcs. Let $C$ be the set of commodities, given as a set of point-to-point demands to transport a certain amount of hazmats. For any commodity $c\in C$, let $s^c$ and $t^c$ be the source node and the destination node respectively, and let $V^c$ be the number of available trucks for the transportation of commodity $c$. Each commodity is associated with a unique type of hazmat. We assume that the risk is computed on each arc of the network and is proportional to the number of trucks traversing such an arc. Although such a proportionality is not always given in pracice, our assumption is a good approximation if the trucks are of the same type and/or the load is very similar as it is the case for standardized castor containers for radioactive material\footnote{Note that changing the proposed model towards different types of trucks with different types of loads and therefore different resulting risks is straightforward and is expected to cause no additional difficulties for the proposed evolutionary algorithm.}. With each arc $(i,j) \in A$, we also associate a cost $c_{ij}$, induced by the travel of a truck on this arc.

@@

The first objective in (\ref{eq:obj1}) is a cost function and is to be minimized. The second objective, given by (\ref{eq:obj2}), minimizes the total risk over all arcs and objective (\ref{eq:obj3}) minimizes the maximum risk imposed on all arcs. The constraints in (\ref{eq:flow}) are conservation constraints, where $\delta^-(i) = \{j \in N: (j,i)\in A\}$ and $\delta^+(i) = \{j \in N: (i,j)\in A\}$ are the direct successors and predecessors of node $i$, and $q_i^c=V^c$ if  $i=s^c$, $q_i^c=-V^c$ if  $i=t^c$ and $q_i^c=0$ otherwise. Note that while the first two objectives are linear in the variables, the maximum in the third objective makes it highly non-linear---the main reason why we propose to use an evolutionary algorithm to deal with the above hazmat routing problem.

@@

\subsection{Representation}
We choose a variable length representation as it has been theoretically shown to be a good choice for multiobjective shortest paths problems \citep{horo2009a}: A solution is thereby represented by a list of paths of variable lengths with one path per truck. We decided to consider a fixed amount of trucks for each commodity and therefore a fixed number of paths through the network although, in principle, the number of trucks could change during the algorithm run if an appropriate mutation operator is designed. The left hand side of Fig.~\ref{fig:paths} shows an exemplary solution with two trucks for commodity 1 and one truck for commodity 2.

@@

\subsubsection{A Simple Random Initialization}
Based on the idea of representing a solution by the beginning of the truck paths which are completed by the shortest path to the destination nodes, we propose the following random initialization:
\begin{enumerate}
	\item Choose for each truck a node in the network uniformly at random among all nodes which are not the source.
	\item Initialize the truck paths with the shortest path from the associated source node to the randomly selected node.
\end{enumerate}
Note that this initialization scheme, though random, does not produce fully random paths. Instead, the initialization results in truck paths which are constructed by concatenating two shortest paths---the shortest path from the source node to the one randomly selected and the shortest path from the random node to the destination, which is added due to the representation described above.

\subsubsection{A Cost-Optimal Initialization}
The chosen representation allows us another initialization which starts each algorithm run from a population consisting of copies of a single cost-optimal solution. To this end, we initialize the paths $p$ for all trucks as empty, containing only the corresponding source node ($p=(s^c)$). With the above proposed representation, this corresponds to the situation where all trucks choose the shortest $s^c$--$t^c$ path for their assigned commodity---implying the smallest possible overall cost but a possibly high risk along the used route(s). Nevertheless, since no other solution can have a smaller cost, the initial solution is already (weakly) Pareto-optimal and is expected to be a good starting point for the algorithm. In the following experiments, we will see when each of the two initialization schemes is favourable.

@@

After the initialization of the population with $\alpha$ individuals by the variator, their evaluation and sending the objective values to the selector module, a standard PISA run always continues in the same way. The selector decides which $\mu$ solutions to variate and the variator uses those $\mu$ solutions as starting points for the generation of $\lambda$ new search points which are again evaluated and send to the selector. In addition, the selector can decide whether solutions are stored in an archive or whether they are deleted. In the case of the selectors used in this study, no archive is maintained. The search stops after a predefined number of variator/selector iterations, also called generations, are performed. In our case, $\alpha=\mu=\lambda$, which will be also called population size hereafter. For more details about PISA, we refer to \citep{bltz2003a}.

\subsection{Hypervolume Estimation Algorithm for Multiobjective Optimization: The HypE Selector of PISA}
In order to fully describe the proposed EMO algorithm for the multiobjective risk-equity constrained routing problem, we briefly mention the working principles of the HypE selection. The main purpose of an EMO algorithm's environmental selection is to ensure convergence of the population towards the Pareto front as well as a good spread of it along the front \citep{deb2001a}. HypE persues these two goals by a two-step approach based on the hypervolume indicator \citep[see][]{ztlf2003a}.

Given the $\mu$ solutions of the old population and the $\lambda$ new ones, the best $\mu$ of them are chosen in the following way. First, a non-dominated sorting \citep{gold1989a} is performed---identifying sets of non-dominated fronts with different ranks. The solutions that are not dominated in the entire set get a rank of 1, the solutions of rank 2 are those which are non-dominated among the remaining solutions, and so forth. As long as the number of solutions is not exceeding $\mu$, entire fronts are chosen for the next iteration starting from the front with rank 1. For the first front for which the overall number of solutions exceeds $\mu$, a second criterion is used to determine the solutions to be kept for the next iteration. This decision is made according to the (expected) loss in hypervolume, which is a criterion that is refining the underlying Pareto dominance relation \citep{ztb2010a}---the main reason why many recent EMO algorithms employ the hypervolume indicator.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Experimental analysis}\label{sec:experiments}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In the remainder of this paper, we investigate how the above proposed evolutionary multiobjective optimization algorithm can cope with realistic instances of the multiobjective risk-equity constrained routing problem. The experimental analyses have two main goals. On the one hand, we show that the algorithm is efficient and can find multiple non-dominated solutions on practical examples in reasonable time. On the other hand, we investigate which of the two proposed initialization schemes is favourable and when this is the case. In addition, a few more experiments are conducted to show that the standard parameters of the algorithm behave as with other test problems. To this end, we vary the population size of the algorithm as well as its environmental selection scheme.

As a test case for all the experiments, we chose to use a real-world case study based on the work of \citet{bcg2009a}. The network used here consists of 311 nodes and 441 arcs and represents the main transport roads in Italy's Lazio region which, amongst other cities, includes the capital Rome. We distinguish between instances with a different number of commodities (between 2 and 4) and with different numbers of trucks, see Table~\ref{tab:instances}. While the costs $c_{ij}$ are proportional to the real distance between the two locations $i$ and $j$, the risks $r_{ij}^c$ for each arc $(i,j) \in A$ and commodity $c \in C$ have been computed as a product of the arc's length, the actual population density in the area of the arc and a constant which depends on the commodity, for details see \citep{bcg2009a} and Table~\ref{tab:instances}. Solely one instance is created differently: to show the impact of correlation between the costs and the risks, we created the instance uncorrelated\_4 with the same graph and risks as for the instance lazio\_4\_1 but where the costs of each arc are the reciprocal of the risk of the first commodity. Figure~\ref{fig:correlation} shows those differences between the costs and risks for the instances lazio\_4\_1 and uncorrelated\_4.

@@

\subsection{Differences Between Instances}
The basis for our experiments are 5 instances of the lazio network with 2 (lazio\_2), 3 (lazio\_3), and 4 commodities (lazio\_4\_1, lazio\_4\_2, and lazio\_4\_3) which differ also in the number of trucks for the commodities. Whereas the number of trucks for lazio\_3 and lazio\_4\_1 have been chosen randomly, lazio\_2 has the same number of trucks than the lazio\_3 instance for the first two commodities. The lazio\_4\_2 instance has only 1 truck more for commodity 4 than lazio\_4\_1 and overall three times the number of trucks for each commodity than the instance lazio\_4\_3.

Figure~\ref{fig:compareAll} shows the non-dominated solutions of 10 independent HypE runs for each instance. What can be observed are three main statements. First of all, cost and risks of the found non-dominated solutions are highly correlated, as they are for each single arc in the problem formulations; therefore, the problem is not truly three-objective but rather bi-objective. Secondly, the range of the objective function values for the found non-dominated solutions follow the number of trucks in the problem instance, i.e., the more trucks a problem instance has the more costs, total and equity risk can be expected in the solutions found. However and thirdly, this relationship is not linear in the plots of Fig.~\ref{fig:compareAll} as can be seen, for example, between the lazio\_4\_2 and the lazio\_4\_3 instance. Though the number of trucks per commodity differs by a factor of 3 between them, the objective function values of the resulting non-dominated solutions after 1,000 generations are slightly further away. The reason for that is the much smaller search space for the lazio\_4\_3 instance which can be searched more exhaustively by the algorithm with the same number of function evalutions---resulting in a better convergence towards the Pareto front.

@@

\subsection{Convergence Towards the Pareto Front}
Figure~\ref{fig:hypervolumeOverTime} shows the boxplots of the hypervolume values for HypE with population size 50 and 100 for different number of generations. What can be seen is that, in general, the hypervolume indicator is mazimized over time and that for the algorithm with population size 50, the hypervolume value starts to stagnate starting from around generation 1500 with the mean still going up but the variance staying high. The fact that all runs found a similar hypervolume value which is significantly higher than the one of the random population in the beginning (around $9.62\cdot 10^26$ for both algorithms) indicates that the algorithm found solution sets which are close to the Pareto front.

@@

\subsection{Random Initialization Versus Cost-Optimal Initialization}
In order to see the impact of the two different initialization schemes \emph{random} and \emph{cost-optimal}, we performed 10 runs of HypE with both initialization schemes for 2,000 generations and a population size of 50 as above. Figure~\ref{fig:randVsCostOriginal} shows the resulting non-dominated solutions of the final populations. Surprisingly, the cost-optimal initialization could not produce other solutions than the cost-optimal one itself. As we will see below, the reason is a high correlation between the risk and costs of each arc, which seemingly result in the fact that no non-dominated solution can be reached from the initial solution by the applied mutation operator. However, the comparison between the both initialization schemes on the lazio\_4\_1 instance shows that also the algorithm employing the random initialization can find solutions which are close to the Pareto front---improving over the hypervolume of the single cost-optimal solution in later generations, see the left-hand plot of Fig.~\ref{fig:costVsRandBoxPlots}.

To show that the cost-optimal initialization is not useless, we run the algorithms again on the instance uncorrelated\_4 where the cost and risk values are fully uncorrelated. In this case, there are more local changes of the truck paths that result in non-dominated solutions and eventually in a larger variaty of solutions found by the algorithm employing the cost-optimal initialization (see Fig.~\ref{fig:randVsCostNew}). Due to the fact that the cost-optimal initialization starts already on the Pareto-front, it also results in higher hypervolume values than for the random initialization variant---both for a low and a high number of generations, see the right-hand side plot of Fig.~\ref{fig:costVsRandBoxPlots}.

@@

\subsection{Changing the Population Size}
In the next experiment, we aim at showing that changing the population size of the algorithm in standard ranges does not change the algorithm outcome drastically. To this end, Fig.~\ref{fig:popsizes} shows the nondominated solutions of 10 independent runs after 100,000 function evaluations for three different versions of HypE where the population size changes from 50 over 100 to 200 on the largest instance lazio\_4\_1 with 4 commodities. From the visual inspection of the resulting solutions, we cannot make a statement about which population size to favour. When comparing the hypervolume values of the resulting populations, however, an effect is visible. Figure~\ref{fig:algoComparison} shows the corresponding box plots of the hypervolume values and one-sided Wilcoxon rank sum tests on the populations' hypervolume indicator values reveil statistical significant differences between two of the three possible algorithm pairs. With a $p$-value of $0.02163$, the null hypothesis of no differences between a population size of 50 and a population size of 100 has to be rejected in favor of stating that the algorithm variant with population size 50 is superior to the one with population size 100. For the comparison between population size 50 and population size 200, the $p$-value equals $0.0144$ with the population size 50 being better whereas the comparison between population size 100 and population size 200 does not reveal a statistical difference (the $p$-values of the one-sided Wilcoxon tests are $0.2179$ and $0.8035$ respectively).

Although these results indicate that smaller population sizes might be always favorable, this is not the case as additional experiments with a population size of 25 show (see the box plot in Fig.~\ref{fig:algoComparison}). The Wilcoxon rank sum test comparing population sizes 25 and 50 results in a $p$-value of $0.004465$ whereas the other comparisons are not statistically significantly different. To conclude, among the tested population sizes, 50 seems to be the best choice for budgets around 100,000 function evaluations. Also with longer runs, we recommend to keep the population size of the algorithm around 50 as the HypE implementation in PISA becomes too costly if the population size is much larger. Ten runs with 2000 iterations and population size 200, for example, need already around 20 hours to complete on an Intel Core2 Duo with 2.8GHz.

@@

Regarding the two initialization schemes, it turned out that in combination with the (local and quite simple) variation operator, the cost-optimal initialization does not allow the algorithm to find additional solutions besides the initial one. However, if the cost and total risk objectives are less correlated, this cost-optimal initialization has advantages over the random initialization in the early stage of the search when compared with respect to the hypervolume indicator. In general, however, we suggest to use the random initialization if the number of function evaluations is not a limiting factor.

>>>>

\begin{abstract}
Nowadays, various hazardous materials, or \emph{hazmats}, are transported on our streets, via trains, planes or by ships. In particular when routing them over the street network, where typically several possible discrete routes exist, one would like to minimize both the costs and the total risks. However, sometimes one aims at minimizing an additional type of risk, which in most optimization models is resulting in a non-linear objective: the equity risk, or the \emph{maximum} risk over all regions, streets, etc. In other words, one aims at distributing the overall risks as much as possible while the costs and the total risk should be still minimized simultaneously. In this paper, this three-objective optimization problem is tackled with a novel randomized algorithm, based on a state-of-the-art Evolutionary Multiobjective Optimization algorithm and tailored to the needs of finding Pareto-optimal truck routes in a given network by proposing a variable-length representation and problem-specific operators. Experimental results on realistic instances for the street network of Italy's Lazio region show that the proposed approach efficiently finds a variety of good solutions in reasonable time. The experiments also show the influence of the algorithm's different building blocks and the instances' parameters on the solution quality.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The transportation of hazardous materials ({\em hazmat} from now on) has received a large interest in recent years, which results from the increase in public awareness of the dangers of hazmats and the enormous amount of hazmats being transported \citep{etv2007a,cd2008a}. The main target of this problem is to select routes for a given origin-destination pair of nodes such that the risk for the surrounding population and the environment is minimized when the hazmat is transported over the selected routes---without producing excessive economic costs.

More specifically, one can distinguish between two main problem types of different scale \citep{etv2007a}. The \emph{local} hazmat routing problem aims at finding the route for a single shipment from a source to a destination node in a given network; the \emph{global} hazmat routing problem, on the other hand, looks at the macro level of shipping several hazmat types simultaneously from multiple sources to multiple destinations---with much less studies on the latter \citep{etv2007a}.

When solving a global hazmat routing problem by minimizing both cost and the total risk, typically several vehicles share the same (short) routes which results in high risks associated to regions surrounding these paths whereas other regions might not be affected at all. In this case, one may wish to distribute the risk in an equitable way over the population and the environment. Although several studies consider this additional minimization of the equity risk, most of them restrict the model to a \emph{single} origin-destination hazmat routing for a specific hazmat, transport mode and vehicle type (see for example \cite{aeb2000a}).
More realistic \emph{multi}-commodity flow model have been proposed where each commodity is considered as one hazmat type \citep{cd2008a} but the majority of all hazmat routing studies do not consider the multi-objective nature of the hazmat routing problem but optimize only the risk \cite{gkbk1990a} or an aggregation or scalarization of the objective functions instead \citep{cd2008a}. An exception is \citep{dgs2005a} where the focus lies on finding a set of dissimilar paths that are Pareto-optimal with respect to cost and risk but the multiobjective hazmat routing problem is only solved indirectly. In addition, also only one origin-destination pair is considered in \citep{dgs2005a}.

As it is important for a decision maker to study the trade-offs among the objectives, we treat the global hazmat routing problem of finding a set of routes minimizing total costs, total risk, and equity risk, from a true multiobjective perspective. Evolutionary Multiobjective Optimization (EMO) algorithms have been shown to be able to compute a set of solutions showing those trade-offs between multiple objectives within a single algorithm run in particular when other (exact) algorithms have difficulties to approach the Pareto front, e.g, if the front is too large \citep{deb2001a,cvl2007a}. As general black-box algorithms, they also allow to cope with changes in the problem formulation which occurs frequently in real-world applications.

In this study, we therefore propose an EMO algorithm for the global hazmat routing problem (Sec.~\ref{S_MEO}). Before, we formalize the specific multiobjective risk-equity constrained routing problem with three objectives (minimize total routing cost, total routing risk {\em and} risk equity) based on the multicommodity flow model of \cite{cd2008a} since this model is the most realistic one permitting to manage several hazmat types simultaneously (Sec.~\ref{S_RECRP}). Section~\ref{sec:experiments} is then devoted to the experimental analysis of this algorithm and its application to realistic instances in the Lazio region of Italy before Sec.~\ref{S_FW} concludes the paper.

@@

Let the transportation network be represented as a directed graph $G = (N, A)$, with $N$ being the set of nodes and $A$ the set of arcs. Let $C$ be the set of commodities, given as a set of point-to-point demands to transport a certain amount of hazmats. For any commodity $c\in C$, let $s^c$ and $t^c$ be the source node and the destination node respectively, and let $V^c$ be the number of available trucks for the transportation of commodity $c$. Each commodity is associated with a unique type of hazmat. We assume that the risk is computed on each arc of the network and is proportional to the number of trucks traversing such an arc. Although such a proportionality is not always given in practice, our assumption is a good approximation if the trucks are of the same type and/or the load is very similar as it is the case for standardized castor containers for radioactive material\footnote{Note that changing the proposed model towards different types of trucks with different types of loads and therefore different resulting risks is straightforward and is expected to cause no additional difficulties for the proposed EMO algorithm.}. With each arc $(i,j) \in A$, we also associate a cost $c_{ij}$, induced by the travel of a truck on this arc.

@@

The first objective in (\ref{eq:obj1}) is the cost function and is to be minimized. The second objective, given by (\ref{eq:obj2}), minimizes the total risk over all arcs and objective (\ref{eq:obj3}) minimizes the maximum risk imposed on all arcs. The constraints in (\ref{eq:flow}) are conservation constraints, where $\delta^-(i) = \{j \in N: (j,i)\in A\}$ and $\delta^+(i) = \{j \in N: (i,j)\in A\}$ are the direct successors and predecessors of node $i$, and $q_i^c=V^c$ if  $i=s^c$, $q_i^c=-V^c$ if  $i=t^c$ and $q_i^c=0$ otherwise. Note that while the first two objectives are linear in the variables, the maximum in the third objective makes it highly non-linear---another reason to use an EMO algorithm to deal with the above hazmat routing problem.

@@

\subsection{Representation}
We choose a variable length representation as it has been theoretically shown to be a good choice for multiobjective shortest paths problems \citep{horo2009a}: A solution is thereby represented by a list of paths of variable lengths with one path per truck. We decided to consider a fixed amount of trucks for each commodity and therefore a fixed number of paths through the network, although, in principle, the number of trucks could change during the algorithm run if an appropriate mutation operator is designed. The left hand side of Fig.~\ref{fig:paths} shows an exemplary solution with two trucks for commodity 1 and one truck for commodity 2.

@@

\subsubsection{A Simple Random Initialization}
Based on the idea of representing a solution by the beginning of the truck paths which are completed by the shortest path to the destination nodes, we propose the following random initialization:
\begin{enumerate}
	\item Choose for each truck a node in the network uniformly at random among all nodes which are not the source.
	\item Initialize the truck paths with the shortest path from the associated source node to the randomly selected node.
\end{enumerate}
Note that this initialization scheme, though random, does not produce fully random paths. Instead, the initialization creates truck paths which are constructed by concatenating two shortest paths---the shortest path from the source node to the one randomly selected and the shortest path from the random node to the destination, which is added due to the representation described above.

\subsubsection{A Cost-Optimal Initialization}
The chosen representation allows us another initialization which starts each algorithm run from a population consisting of copies of a single cost-optimal solution. To this end, we initialize the paths $p$ for all trucks as empty, containing only the corresponding source node ($p=(s^c)$). With the above proposed representation, this corresponds to the situation where all trucks choose the shortest $s^c$--$t^c$ path for their assigned commodity---implying the smallest possible overall cost but a possibly high risk along the used route(s). Nevertheless, since no other solution can have a smaller cost, the initial solution is already (weakly) Pareto-optimal and is expected to be a good starting point for the algorithm. In the following experiments, we will see when each of the two initialization schemes is favorable.

@@

After the initialization of the population with $\alpha$ solutions by the variator, the new solutions are evaluated and their objective vectors send to the selector module. Afterwards, a PISA run always continues in the same way. The selector decides which $\mu$ solutions to variate and the variator uses those $\mu$ solutions as starting points for the generation of $\lambda$ new search points which are again evaluated and send to the selector. In addition, the selector can decide whether solutions are stored in an archive or whether they are deleted. In the case of the selectors, used in this study, no archive is maintained. The search stops after a predefined number of variator/selector iterations, also called generations. In our case, $\alpha=\mu=\lambda$, which will be also called population size hereafter. For more details about PISA, we refer to \citep{bltz2003a}.

\subsection{Hypervolume Estimation Algorithm for Multiobjective Optimization: The HypE Selector of PISA}
In order to fully describe the proposed EMO algorithm for the multiobjective risk-equity constrained routing problem, we briefly mention the working principles of the HypE selection. The main purpose of an EMO algorithm's environmental selection is to ensure convergence of the population towards the Pareto front as well as a good spread of it along the front \citep{deb2001a}. HypE pursues these two goals by a two-step approach based on the hypervolume indicator (see \citep{ztlf2003a} for a definition of the hypervolume indicator).

Given the $\mu$ solutions of the old population and the $\lambda$ new ones, the best $\mu$ of them are chosen in the following way. First, a non-dominated sorting \citep{gold1989a} is performed---identifying sets of non-dominated fronts with different ranks. The solutions that are not dominated in the entire set get a rank of 1, the solutions of rank 2 are those which are non-dominated among the remaining solutions, and so forth. As long as the number of solutions is not exceeding $\mu$, entire fronts are chosen to be introduced in the next generation's population; starting from the front with rank 1. For the first front for which the overall number of solutions exceeds $\mu$, a second criterion is used to determine the solutions to be kept for the next iteration. This decision is made according to the (expected) loss in hypervolume, which is a criterion that is refining the underlying Pareto dominance relation \citep{ztb2010a}---the main reason why many recent EMO algorithms employ the hypervolume indicator.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Experimental analysis}\label{sec:experiments}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In the remainder of this paper, we investigate how the above proposed EMO algorithm can cope with realistic instances of the multiobjective risk-equity constrained routing problem. The experimental analyses have two main goals. On the one hand, we show that the algorithm is efficient and can find multiple non-dominated solutions on practical examples in reasonable time. On the other hand, we investigate which of the two proposed initialization schemes is favorable and when this is the case. In addition, a few more experiments are conducted to show that the standard parameters of the algorithm behave as with other test problems. To this end, we vary the population size of the algorithm as well as its environmental selection scheme.

As a test case for all the experiments, we choose to use a real-world case study based on the work of \citet{bcg2009a}. The network used here consists of 311 nodes and 441 arcs and represents the main transport roads in Italy's Lazio region which, amongst other cities, includes the capital Rome. We distinguish between instances with a different number of commodities (between 2 and 4) and with different numbers of trucks, see Table~\ref{tab:instances}. While the costs $c_{ij}$ are proportional to the real distance between the two locations $i$ and $j$, the risks $r_{ij}^c$ for each arc $(i,j) \in A$ and commodity $c \in C$ have been computed as a product of the arc's length, the actual population density in the area of the arc and a constant which depends on the commodity, for details see \citep{bcg2009a} and Table~\ref{tab:instances}. Solely one instance is created differently: to show the impact of correlation between the costs and the risks, we created the instance uncorrelated\_4 with the same graph and risks as for the instance lazio\_4\_1 but where the costs of each arc are the reciprocal of the risk of the first commodity. Figure~\ref{fig:correlation} shows those differences between the costs and risks for the instances lazio\_4\_1 and uncorrelated\_4.

@@

\subsection{Differences Between Instances}
The basis for our experiments are 5 instances of the lazio network with 2 (lazio\_2), 3 (lazio\_3), and 4 commodities (lazio\_4\_1, lazio\_4\_2, and lazio\_4\_3) which differ also in the number of trucks for the commodities. Whereas the number of trucks for lazio\_3 and lazio\_4\_1 have been chosen randomly, lazio\_2 has the same number of trucks than the lazio\_3 instance for the first two commodities. The lazio\_4\_2 instance has only 1 truck more for commodity 4 than lazio\_4\_1 and overall three times the number of trucks for each commodity than the instance lazio\_4\_3, cp.\ Table~\ref{tab:instances}.

Figure~\ref{fig:compareAll} shows the non-dominated solutions of 10 independent HypE runs for each instance. What can be observed are three main statements. First of all, cost and risks of the found non-dominated solutions are highly correlated, as they are for each single arc in the problem formulations; therefore, the problem is not truly three-objective but rather bi-objective. Secondly, the range of the objective function values for the found non-dominated solutions follow the number of trucks in the problem instance, i.e., the more trucks a problem instance has the more costs, total and equity risk can be expected in the solutions found. However and thirdly, this relationship is not linear in the plots of Fig.~\ref{fig:compareAll} as can be seen, for example, between the lazio\_4\_2 and the lazio\_4\_3 instance. Though the number of trucks per commodity differs by a factor of 3 between them, the objective function values of the resulting non-dominated solutions after 2,000 generations are slightly further away. The reason for that is the much smaller search space for the lazio\_4\_3 instance which can be searched more exhaustively by the algorithm with the same number of function evaluations---resulting in a better convergence towards the Pareto front.

@@

\subsection{Convergence Towards the Pareto Front}
Figure~\ref{fig:hypervolumeOverTime} shows the boxplots of the hypervolume values for HypE with population size 50 and 100 for different number of generations. What can be seen is that, in general, the hypervolume indicator is maximized over time and that for the algorithm with population size 50, the hypervolume value starts to stagnate starting from around generation 1500 with the mean still going up but the variance staying high. The fact that all runs found a similar hypervolume value which is significantly higher than the one of the random population in the beginning (around $9.62\cdot 10^{26}$ for both algorithms) indicates that the algorithm found solution sets which are close to the Pareto front.

@@

\subsection{Random Initialization Versus Cost-Optimal Initialization}
In order to see the impact of the two different initialization schemes \emph{random} and \emph{cost-optimal}, we performed 10 runs of HypE with both initialization schemes for 2,000 generations and a population size of 50 as above. Figure~\ref{fig:randVsCostOriginal} shows the resulting non-dominated solutions of the final populations. Surprisingly, the cost-optimal initialization could not produce other solutions than the cost-optimal one itself. As we will see below, the reason is a high correlation between the risk and costs of each arc, which seemingly results in the fact that no non-dominated solution can be reached from the initial solution by the applied mutation operator. However, the comparison between the two initialization schemes on the lazio\_4\_1 instance shows that also the algorithm employing the random initialization can find solutions which are close to the Pareto front---improving over the hypervolume of the single cost-optimal solution in later generations, see the left-hand plot of Fig.~\ref{fig:costVsRandBoxPlots}.

To show that the cost-optimal initialization is not useless, we run the algorithms again on the instance uncorrelated\_4 where the cost and risk values are fully uncorrelated. In this case, there are more local changes of the truck paths that result in non-dominated solutions and eventually in a larger variety of solutions found by the algorithm employing the cost-optimal initialization (see Fig.~\ref{fig:randVsCostNew}). Due to the fact that the cost-optimal initialization starts already on the Pareto-front, it also results in higher hypervolume values than for the random initialization variant---both for a low and a high number of generations, see the right-hand side plot of Fig.~\ref{fig:costVsRandBoxPlots}. All algorithms of Fig.~\ref{fig:costVsRandBoxPlots} are pairwisely significantly different with respect to their hypervolume values according to Wilcoxon rank sum tests with p-values below 0.01.

@@

\subsection{Changing the Population Size}
In the next experiment, we aim at showing that changing the population size of the algorithm in standard ranges does not change the algorithm outcome drastically. To this end, Fig.~\ref{fig:popsizes} shows the nondominated solutions of 10 independent runs after 100,000 function evaluations for three different versions of HypE where the population size changes from 50 over 100 to 200 on the largest instance lazio\_4\_1 with 4 commodities. From the visual inspection of the resulting solutions, we cannot make a statement about which population size to favor. When comparing the hypervolume values of the resulting populations, however, an effect is visible. Figure~\ref{fig:algoComparison} shows the corresponding box plots of the hypervolume values and one-sided Wilcoxon rank sum tests on the populations' hypervolume indicator values reveal statistical significant differences between two of the three possible algorithm pairs. With a $p$-value of $0.02163$, the null hypothesis of no differences between a population size of 50 and a population size of 100 has to be rejected in favor of stating that the algorithm variant with population size 50 is superior to the one with population size 100. For the comparison between population size 50 and population size 200, the $p$-value equals $0.0144$ with the population size 50 being better whereas the comparison between population size 100 and population size 200 does not reveal a statistical difference (the $p$-values of the one-sided Wilcoxon tests are $0.2179$ and $0.8035$ respectively).

Although these results indicate that smaller population sizes might be always favorable, this is not the case as additional experiments with a population size of 25 show (see the box plot in Fig.~\ref{fig:algoComparison}). The Wilcoxon rank sum test comparing population sizes 25 and 50 results in a $p$-value of $0.004465$ whereas the other comparisons are not statistically significantly different. To conclude, among the tested population sizes, 50 seems to be the best choice for budgets around 100,000 function evaluations but the population size is not a parameter that is difficult to adjust for the user. According to the runtime of the algorithm, we recommend to keep the population size of the algorithm around 50 as the HypE implementation in PISA becomes too costly if the population size is much larger. Ten runs with 2000 iterations and population size 200, for example, need already around 20 hours to complete on an Intel Core2 Duo with 2.8GHz.

@@

Regarding the two initialization schemes, it turns out that in combination with the (local and quite simple) variation operator, the cost-optimal initialization does not allow the algorithm to find additional solutions besides the initial one on the Lazio instances. However, if the cost and total risk objectives are uncorrelated, this cost-optimal initialization has advantages over the random initialization in all stages of the search when compared with respect to the hypervolume indicator. In general, however, we suggest to use the random initialization\footnote{or a combination with a more exploring mutation operator} if the costs and risks are highly correlated.
